---
title: 'Practicum 9: Visualization'
author: 'Data Science for Biomedical Informatics (BMIN503/EPID600)'
output: 
  html_document:
    toc: true
    toc_float: 
        collapsed: true
        smooth_scroll: true
    depth: 3 
    theme: paper 
    highlight: tango
---

```{r global options, include=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE)
```   
***
New packages to install for this practicum:
```{r, eval=FALSE}
install.packages("devtools")
library(devtools)
install_github("wilkelab/cowplot")
```
***
Good visualization is critical to present research findings in an intuitive and appealing way, but it can also enhance our ability to understand data as we analyze it and uncover unexpected patterns. In this practicum, we will learn more about the `ggplot2` package, which has greatly simplified the production of high quality graphics in R. The official [`ggplot2`](http://docs.ggplot2.org/current/) documentation is helpful to give an idea of how much this package can do. You can also read more in this [overview](http://ggplot2.tidyverse.org/) and/or the [data visualization](http://r4ds.had.co.nz/data-visualisation.html) chapter of the "R for Data Science" book. Samples of plots that can be made with `ggplot2` with accompanying code are available at sites such as [UBC Stat shiny server](http://shiny.stat.ubc.ca/r-graph-catalog/). We will use (1) `iris`, a classic R built-in dataset that contains measures of sepal length and width, and petal length and width for 50 flowers from each of 3 species of iris (setosa, versicolor, virginica), and (2) the NHANES dataset we've used before.

```{r eval=TRUE, message=FALSE}
library(dplyr)
library(ggplot2)
nhanes <- read.csv(url("https://raw.githubusercontent.com/HimesGroup/BMIN503/master/DataFiles/NHANES_2007to2008.csv"), header=TRUE)
nhanes <- rename(nhanes, id=SEQN, gender=RIAGENDR, age=RIDAGEYR, race=RIDRETH1, education=DMDEDUC2, income=INDHHIN2, health.provider=HUQ040, wheezing=RDQ070, asthma=MCQ010, voc=WTSVOC2Y, fvc=SPXNFVC, fev1=SPXNFEV1) %>%
    mutate(gender=factor(gender, levels=c(1, 2), labels=c("male", "female"))) %>%
    mutate(race=factor(race, levels=c(3, 1, 2, 4, 5), labels=c("white", "MexicanAmerican", "Hispanic", "black", "other"))) %>%
    mutate(asthma=ifelse(asthma %in% c(7,9), NA, ifelse(asthma==2, 0, 1))) %>%
    filter(!is.na(asthma))
str(nhanes)
str(iris)
```

### Pipe operator works with ggplot
```{r eval=TRUE, message=FALSE}
nhanes %>% 
    mutate(adult.status = factor(age>=18, labels=c("child", "adult"))) %>% 
    filter(adult.status=="child") %>% 
    ggplot(aes(x=age, y=fev1)) + 
        geom_point() +
        geom_smooth(method = "lm", color="red")
```


### More ggplot options and plot types
We'll look at more options related to ggplot by adding to the layers created before. More options related to setting background, axis, title, and legend options are available in [`theme`](http://ggplot2.tidyverse.org/reference/theme.html). The default settings for ggplot will appropriately select color type (although not necessary color blind or black-and-white printer friendly).

#### Scatterplot
```{r eval=TRUE}
ggplot(data=iris, aes(Sepal.Length, Petal.Length, color=Petal.Width)) + 
    geom_point() + 
    labs(title="Iris Characteristics") +
    labs(x="Sepal Length", y="Petal Length") + #Adds a layer with labels
    xlim(c(4, 8)) + #Changes x axis limits
    ylim(c(1, 8))   #Changes y axis limits
```

Change color scale.
```{r eval=TRUE}
ggplot(data=iris, aes(Sepal.Length, Petal.Length, color=Petal.Width)) + 
    geom_point() + 
    scale_color_gradient(low="#fee6ce", high="#e6550d") +
    labs(title="Iris Characteristics") +
    labs(x="Sepal Length", y="Petal Length") + #Adds a layer with labels
    xlim(c(4, 8)) + #Changes x axis limits
    ylim(c(1, 8)) + #Changes y axis limits
    theme_bw() #Changes theme to black and white
```

This time we can view up to four dimensions of our data by setting color and size, in addition to x and y variables.
```{r eval=TRUE}
ggplot(data=iris, aes(Sepal.Length, Petal.Length, color=Species, size=Petal.Width)) + 
    geom_point() + 
    labs(title="Iris Characteristics") +
    labs(x="Sepal Length", y="Petal Length") + #Adds a layer with labels
    xlim(c(4, 8)) + #Changes x axis limits
    ylim(c(1, 8)) + #Changes y axis limits
    theme_bw() #Changes theme to black and white
```
  
#### Barplot
When creating barplots, use `geom_bar` with default option `stat=bin` to get counts in each category. To split panels according to another categorical variable `facet_grid` can be used. More color options that work for all chart types are described [here](http://www.cookbook-r.com/Graphs/Colors_(ggplot2)/).
```{r eval=TRUE}
ggplot(data=nhanes, aes(x=race, fill=factor(asthma))) + 
    geom_bar(position="dodge")
ggplot(data=nhanes, aes(x=race, fill=factor(asthma))) + 
    geom_bar(position="dodge") +
    facet_grid(. ~gender) + #Split by another variable
    theme(axis.text.x=element_text(angle=45, hjust=1))
ggplot(data=nhanes, aes(x=gender, fill=factor(asthma))) + 
    geom_bar() +
    scale_fill_brewer(palette="Set1") #ColorBrewer palettes
```

#### Boxplot
```{r eval=TRUE}
ggplot(data=iris, aes(Species, Sepal.Length)) + 
    geom_boxplot() 
ggplot(data=iris, aes(Species, Sepal.Length)) +
    geom_boxplot() +
    geom_jitter() #Add a second layer of jittered data points
ggplot(data=iris, aes(Species, Sepal.Length)) +
    geom_boxplot(outlier.colour="red", outlier.size=4) #Making outliers more noticeable
ggplot(data=iris, aes(Species, Sepal.Length)) +
    geom_violin() #violin plots
ggplot(data=iris, aes(Species, Sepal.Length)) +
    geom_violin(fill="lightblue", draw_quantiles = c(0.25, 0.5, 0.75)) +
    geom_jitter(height=0, width=0.1)
ggplot(data=iris, aes(Species, Sepal.Length)) +
    geom_boxplot(aes(fill=Species)) #Coloring according to species
ggplot(data=nhanes, aes(gender, voc)) +
    geom_boxplot(color="darkgreen", fill="darkgreen", alpha=0.5)
ggplot(data=nhanes, aes(gender, voc)) +
    geom_boxplot(aes(fill=factor(asthma))) #Add another variable with fill
```

#### Histograms
```{r eval=TRUE}
ggplot(data=nhanes, aes(age)) + 
    geom_histogram(breaks=seq(0, 80, 1), color="blue", fill="blue", alpha=0.7) 
#Density plot
ggplot(data=nhanes, aes(age)) + 
    geom_histogram(aes(y=..density..), breaks=seq(0, 80, 1), color="blue", fill="blue", alpha=0.7) +
    geom_density(color="red")
```

#### Smoothing lines
When many points are part of a scatter plot, it can be difficult to see patterns that represent major trends. The `geom_smooth` command (similar to `geom_stat`) will plot smooth conditional means. The smoothing function can be specified with the `method` option as  "lm", "glm", "gam", "loess", "nls". 
```{r eval=TRUE}
ggplot(nhanes, aes(x=age, y=fev1)) +
    geom_point(color="grey") +
    geom_smooth(color="red") #Default option is method="auto" which selects loess for <1000 points and gam with formula y~s(x, bs="cs") otherwise

ggplot(nhanes, aes(x=age, y=fev1)) +
    geom_point(color="grey") +
    geom_smooth(color="red") +
    geom_smooth(color="black", method="lm") +
    geom_smooth(color="green", method="loess")
```

Additionally, a `formula` can be given for use in the smoothing function.
```{r eval=TRUE}
x <- 1:10
y <- jitter(x^2)
DF <- data.frame(x, y)

ggplot(DF, aes(x=x, y=y)) + 
    geom_point() +
    geom_smooth(method='lm', aes(color = 'linear'), se=FALSE) +
    geom_smooth(method='lm', formula = y ~ poly(x, 2), aes(color = 'polynomial'), se=FALSE) +
    geom_smooth(method='nls', 
                formula = y ~ a*log(x) + b, 
                method.args = list(start = c(a=1, b=1)),
                aes(color = 'logarithmic'), 
                se=FALSE) +
    geom_smooth(method='nls', 
                formula = y ~ a*exp(b*x), 
                method.args = list(start = c(a=1, b=1)),
                aes(color = 'Exponential'), 
                se=FALSE)
```


#### Contours of 2D density estimates
When many points are in a 2D scatter plot, it can similarly be difficult to tell where the majority of points are. Density plots can help us interpret data relationships when there is overplotting.
```{r eval=TRUE}
ggplot(nhanes, aes(x=fvc/1000, y=fev1/1000)) +
    geom_point(color="grey") +
    stat_density2d(aes(fill=..level.., alpha=..level..), geom="polygon", color="black") +
    scale_fill_continuous() +
    guides(alpha="none")
```


#### Raster of 2D density estimates
If the contours are left out, then `stat_density2d` can be used to produce a raster of the data.
```{r eval=TRUE}
ggplot(nhanes, aes(x=fvc/1000, y=fev1/1000)) +
    stat_density2d(aes(fill=..density..), geom="raster", contour=FALSE) +
    scale_fill_continuous(low="#238443", high="#ffffcc") +
    guides(alpha="none")
```  


#### `viridis` color scheme
This color scheme is a personal favorite for use with heatmaps and other 2D plots that, as of 2018, is part of `ggplot2`.
```{r eval=TRUE}
ggplot(nhanes, aes(x=fvc/1000, y=fev1/1000)) +
    stat_density2d(aes(fill=..density..), geom="raster", contour=FALSE) +
    scale_fill_viridis_c() +
    guides(alpha="none")
``` 


### Saving figures
To save graphical output to a file in R, one opens a graphics device to redirect output of plots to them, and when plots are completed, one _closes_ the device to save the final version. To see all available graphical devices, type `?Devices`. The default device in R is usually an interactive window, like the one we see in RStudio. Here are two examples using common formats.
```{r eval=TRUE}
options("device") #To see the current graphics device
fig <- ggplot(data=iris, aes(Species, Sepal.Length)) +
        geom_boxplot(color="darkblue", fill="blue")
#Save as PDF
pdf("Ex1.pdf", width=6, height=5) #Units are inches
fig
dev.off()
#Save as TIFF
tiff("Ex1.tiff", width=480, height=480, units="px")
fig
dev.off()
```
`ggplot2` offers a simplified function to save graphics output: `ggsave`.
```{r eval=TRUE}
fig
ggsave("Ex2.pdf") #Default plot to save is last plot
ggsave("Ex2.tiff", plot=fig)
```


### Publication graphics
The theme used with `ggplot2` can be customized to suit your personal taste. One researcher created an R package called [`cowplot`](https://cran.r-project.org/web/packages/cowplot/vignettes/introduction.html) as a `ggplot2` add-on to easily create publication-ready graphics.
```{r eval=FALSE}
library(cowplot)
save_plot("Ex3.pdf", fig, base_aspect_ratio=1.3) #Save using the default format of cowplot
```



